---
title: "Longitudinal Analysis of Glycaemic Trajectories in Type 2 Diabetes Mellitus"
subtitle: "Linear Mixed Effects Model (LMM) and Generalised Estimating Equations (GEE) | GET503 Advanced Numerical Data Analysis | DrPH Epidemiology | Universiti Sains Malaysia"
author:
  - name: "Siti Nurhafizah binti Saharudin"
  - name: "Ahmad Zulfahmi bin Sha'ari"
  - name: "Nazirul Munir bin Abu Hassan"
  - name: "Mohd Fittri Fahmi bin Fauzi"
date: today
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 4
    toc-location: left
    toc-title: "Contents"
    number-sections: true
    code-fold: true
    code-tools: true
    fig-width: 9
    fig-height: 5
    fig-dpi: 150
    highlight-style: kate
    df-print: paged
    self-contained: true
    css: T2DM_LMM_GEE_Combined.css
execute:
  warning: false
  message: false
  echo: true
  cache: false
---

```{=html}
<script>
document.addEventListener("DOMContentLoaded", function () {
  const toc = document.getElementById("TOC");
  if (!toc) return;
  const level1Items = toc.querySelectorAll(":scope > ul > li");
  level1Items.forEach(function (li) {
    const subList = li.querySelector(":scope > ul");
    if (!subList) return;
    li.classList.add("toc-open");
    const link = li.querySelector(":scope > a");
    if (!link) return;
    link.style.cursor = "pointer";
    link.addEventListener("click", function (e) {
      e.preventDefault();
      const href = this.getAttribute("href");
      if (li.classList.toggle("toc-open")) {
        if (href) window.location.hash = href;
      }
    });
    const level2Items = subList.querySelectorAll(":scope > li");
    level2Items.forEach(function (li2) {
      const subList2 = li2.querySelector(":scope > ul");
      if (!subList2) return;
      const link2 = li2.querySelector(":scope > a");
      if (!link2) return;
      link2.style.cursor = "pointer";
      link2.addEventListener("click", function (e) {
        e.preventDefault();
        const href2 = this.getAttribute("href");
        if (li2.classList.toggle("toc-open")) {
          if (href2) window.location.hash = href2;
        }
      });
    });
  });
});
</script>
```

---

# Overview & Study Background {.unnumbered}

## Assignment Details {.unnumbered}

| Item | Detail |
|------|--------|
| **Assignment** | GET503 Advanced Numerical Data Analysis |
| **Programme** | Doctor of Public Health (DrPH) Epidemiology |
| **Institution** | Universiti Sains Malaysia |
| **Methods** | Linear Mixed Effects Model (LMM) & Generalised Estimating Equations (GEE) |
| **Outcome** | HbA1c (%), longitudinal, 3 time points |
| **Sample** | 20,000 T2DM patients × 3 waves = 60,000 observations |
| **Data source** | [Kaggle: T2DM Longitudinal Treatment Outcomes](https://www.kaggle.com/datasets/sahilrana034/t2dm-longitudinal-treatment-outcomes) |

---

## Background {.unnumbered}

Type 2 Diabetes Mellitus (T2DM) is a chronic metabolic disease characterised by persistent hyperglycaemia and remains a leading cause of cardiovascular disease, renal failure, and premature mortality worldwide. In Malaysia, T2DM constitutes a growing public health burden, with national prevalence estimates exceeding 18% of the adult population. Glycated haemoglobin (HbA1c), reflecting average blood glucose over the preceding two to three months, is the primary clinical marker used to monitor long-term glycaemic control. Reducing HbA1c through targeted lifestyle and pharmacological interventions is a central goal of T2DM management at both the individual clinical and population levels.

This analysis utilises a longitudinal dataset of **20,000 T2DM patients**, each measured at three structured time points, pre-intervention (T1\_Pre), mid-intervention (T2\_Mid), and post-intervention (T3\_Post), capturing a wide range of patient-level characteristics including demographics, lifestyle behaviours, clinical markers, and intervention type. Because each patient contributes three repeated measurements, standard multiple linear regression (MLR) is inappropriate: observations from the same patient are inherently correlated, violating the independence assumption. Two complementary longitudinal methods address this, the **Linear Mixed Effects Model (LMM)** and **Generalised Estimating Equations (GEE)**, which differ fundamentally in their estimand and interpretation:

- **LMM** models *subject-specific* effects. It directly estimates how each patient's HbA1c changes over time by incorporating patient-level random effects, making it ideal for understanding individual trajectories and quantifying between-patient heterogeneity.
- **GEE** models *population-averaged* effects. It estimates the expected HbA1c change for an average patient in the population, making its coefficients directly interpretable for public health policy and programme evaluation.

Both analyses follow the **REAAAPP** workflow (**R**ead → **E**DA → **A**ssumptions → **A**nalyse → **A**ssess → **P**resent → **P**ublish), ensuring a structured, reproducible, and transparent analytical process.

---

## Research Question & Objectives {.unnumbered}

**Primary Research Question**

Among patients with Type 2 Diabetes Mellitus, do different intervention strengths (Weak, Standard, Intensive) produce different trajectories of HbA1c change over time, after adjusting for demographic, lifestyle, and clinical covariates?

**Primary Objective**

To determine whether intervention strength (Weak, Standard, Intensive) produces significantly different HbA1c trajectories over time, using both LMM (subject-specific estimates) and GEE (population-averaged estimates).

**Secondary Objectives**

1. To quantify the overall change in HbA1c over time, independent of intervention group or patient characteristics (unconditional growth model: LMM).
2. To identify patient-level demographic, lifestyle, and clinical factors independently associated with HbA1c in both the LMM and GEE frameworks.
3. To estimate and compare adjusted mean HbA1c at each time point across intervention groups using estimated marginal means (EMMs), and to determine whether all three intervention groups differ significantly from each other at the study endpoint.
4. To select the optimal working correlation structure for GEE using the Quasi-likelihood under Independence Criterion (QIC) and evaluate structural sensitivity through cross-structure comparison.

---

## Study Variables {.unnumbered}

**Outcome variable**

| Variable | Type | Description |
|---|---|---|
| `hba1c` | Continuous | HbA1c (%), primary measure of glycaemic control; the dependent variable in all models |

**Time variable**

| Variable | Type | Description |
|---|---|---|
| `time_step` | Ordered factor | Study time point: T1\_Pre (baseline), T2\_Mid, T3\_Post |
| `time_num` | Numeric (0/1/2) | Numeric encoding of time, required for the random slope model |

**Primary exposure (intervention)**

| Variable | Type | Levels |
|---|---|---|
| `intervention_strength` | Ordered factor | Weak *(ref)*, Standard, Intensive |

**Demographic covariates**

| Variable | Type | Description |
|---|---|---|
| `age` | Continuous | Age in years |
| `gender` | Factor | Female *(ref)*, Male |

**Lifestyle covariates**

| Variable | Type | Description |
|---|---|---|
| `physical_activity_level` | Ordered factor | Low *(ref)*, Moderate, High |
| `diet_score` | Continuous | Dietary quality score; higher = better |
| `sleep_hours` | Continuous | Average daily sleep (hours) |

**Clinical covariates**

| Variable | Type | Description |
|---|---|---|
| `bmi` | Continuous | Body Mass Index (kg/m²) |
| `medication_adherence` | Ordered factor | Poor *(ref)*, Moderate, Good |
| `comorbidity_level` | Ordered factor | Low *(ref)*, Medium, High |

**Administrative variables**

| Variable | Type | Description |
|---|---|---|
| `patient_id` | Factor | Unique patient identifier; used as the random-intercept grouping variable in LMM and the cluster ID in GEE |

> **Reference level convention:** All ordered factors are set with the worst or least favourable category as the reference. Regression coefficients therefore represent improvements relative to the most clinically unfavourable state. For gender, Female is set as the reference category throughout.

---

# Data Preparation & Exploratory Analysis

This section covers the **R**ead and **E**xplore steps of the REAAAPP framework. All packages are loaded upfront, the shared dataset is prepared once, and exploratory analyses are conducted before any modelling. This sequence ensures that any data quality issues are identified and resolved prior to inference.

## R: Package Setup

All required packages are loaded using `pacman::p_load()`, which installs any missing packages automatically. The combined package set supports data wrangling, EDA, LMM fitting, GEE fitting, post-estimation, and publication-quality visualisation.

```{r setup}
if (!require("pacman")) install.packages("pacman")

pacman::p_load(
  # Data management
  tidyverse, janitor, here,
  # EDA
  skimr, DataExplorer, corrplot,
  # Tables
  gtsummary, gt, kableExtra,
  # Model output
  broom, broom.mixed, broom.helpers,
  # LMM
  lme4, lmerTest, sjPlot, performance, MuMIn,
  # GEE
  geepack,
  # Post-estimation
  emmeans, car, lmtest, MASS,
  # Visualisation
  ggplot2, patchwork, scales,
  # Utilities
  rstatix, purrr
)

# ── Okabe & Ito (2008) colourblind-safe palette ──────────────
cb_intervention <- c(
  "Weak"      = "#E69F00",
  "Standard"  = "#56B4E9",
  "Intensive" = "#0072B2"
)
cb_time <- c(
  "T1_Pre"  = "#56B4E9",
  "T2_Mid"  = "#009E73",
  "T3_Post" = "#0072B2"
)

# ── Shared ggplot2 theme ──────────────────────────────────────
theme_gee <- function(base_size = 12) {
  theme_bw(base_size = base_size) %+replace%
    theme(
      panel.grid.minor  = element_blank(),
      panel.border      = element_rect(colour = "#cdd8e6"),
      plot.title        = element_text(face = "bold", size = base_size + 1,
                                       colour = "#1a1c2e"),
      plot.subtitle     = element_text(colour = "#3d4560", size = base_size - 1),
      legend.position   = "bottom",
      legend.key.size   = unit(0.75, "lines"),
      legend.title      = element_text(face = "bold"),
      strip.background  = element_rect(fill = "#e6f1fb"),
      strip.text        = element_text(face = "bold", colour = "#0b2d5e"),
      axis.title        = element_text(face = "bold", size = base_size - 0.5),
      axis.text         = element_text(colour = "#3d4560")
    )
}
```

## R: Data Loading & Preparation

The cleaned dataset is loaded from `t2dm_clean.csv`. Factor levels are encoded explicitly to ensure correct reference categories and ordering throughout all subsequent models. Gender uses **Female** as the reference category. The dataset is sorted by `patient_id` and `time_num`, a requirement for `geepack` to correctly identify within-cluster observation order.

::: {.callout-important}
## Before Rendering: Set Your Data Path
Update `data_path` in the chunk below to the full path of your `t2dm_clean.csv` file if it is **not** in the same folder as this `.qmd` file. For example:

```
data_path <- "C:/Users/User/Documents/MyProject/t2dm_clean.csv"
```
:::

```{r data-prep}
# ── ✏️  SET THIS PATH to your t2dm_clean.csv location ────────
# If the CSV is in the same folder as this .qmd, leave as-is.
# Otherwise, provide the full file path, e.g.:
#   data_path <- "C:/Users/User/Documents/t2dm_clean.csv"
data_path <- "t2dm_clean.csv"

df <- read_csv(data_path) |>
  mutate(
    patient_id = factor(patient_id),
    # Numeric time index (0, 1, 2) - required for random slope models
    time_num = as.numeric(factor(time_step,
               levels = c("T1_Pre", "T2_Mid", "T3_Post"))) - 1,

    # Ordered factors - reference category is the first (clinically worst) level
    time_step = factor(time_step,
                  levels = c("T1_Pre", "T2_Mid", "T3_Post")),
    intervention_strength = factor(intervention_strength,
                  levels = c("Weak", "Standard", "Intensive")),
    physical_activity_level = factor(physical_activity_level,
                  levels = c("Low", "Moderate", "High")),
    medication_adherence = factor(medication_adherence,
                  levels = c("Poor", "Moderate", "Good")),
    comorbidity_level = factor(comorbidity_level,
                  levels = c("Low", "Medium", "High")),
    gender = factor(gender, levels = c("Female", "Male"))
  ) |>
  # CRITICAL for GEE: data must be sorted by cluster ID then time
  arrange(patient_id, time_num)

cat(sprintf("Observations : %d\n", nrow(df)))
cat(sprintf("Patients     : %d\n", n_distinct(df$patient_id)))
cat(sprintf("Time points  : %d per patient (balanced)\n",
            nrow(df) / n_distinct(df$patient_id)))
cat(sprintf("HbA1c range  : %.2f - %.2f %%\n",
            min(df$hba1c), max(df$hba1c)))
```

```{r eda-glimpse}
cat("=== Dataset Structure (glimpse) ===\n")
glimpse(df)
```

::: {.callout-note}
## Data Structure Confirmation
`glimpse()` confirms the dataset contains **60,000 rows × 14 variables** with no unexpected data types after factor coding. HbA1c is continuous (`dbl`), `time_step` and `intervention_strength` are correctly ordered factors, and `patient_id` is the unique cluster identifier. The dataset is perfectly balanced with exactly 3 observations per patient, and no missing values are present in any analysis variable.
:::

---

## E: Exploratory Data Analysis

### Data Quality Summary

Three complementary checks confirm data integrity before modelling: `skimr::skim()` provides a full distributional profile of all variables; `colSums(is.na())` counts missing values per column; and `DataExplorer::plot_missing()` visualises the missingness pattern.

```{r eda-summary}
skimr::skim(df)
colSums(is.na(df))
DataExplorer::plot_missing(df)
```

### Balance Check

A longitudinal dataset is **balanced** if every subject contributes the same number of observations at the same time points. This check confirms that each of the 20,000 patients appears exactly three times (once per time point), ruling out phantom rows, duplicate entries, or missing visits that would complicate subsequent analysis.

```{r balance-check}
df |> count(time_step)                 # expect 20,000 per time point
df |> count(patient_id) |> summary()  # expect min = max = 3
```

A perfectly balanced panel means no complications from unequal cluster sizes. Both LMM and GEE handle unbalanced data, but confirmed balance rules out data entry issues and ensures the ICCand QIC computations reflect the intended design.

### Spaghetti Plot: Individual Trajectories

```{r fig-spaghetti, fig.cap="Figure 1. HbA1c individual and mean trajectories across three time points (n = 100 randomly sampled patients; red line = group mean). Wide spread of blue lines indicates substantial between-patient heterogeneity in baseline HbA1c, the primary justification for including random effects in LMM."}
#| fig-width: 8
#| fig-height: 5

set.seed(123)
sample_pts <- sample(levels(df$patient_id), 100)

ggplot(df |> filter(patient_id %in% sample_pts),
       aes(x = time_step, y = hba1c, group = patient_id)) +
  geom_line(alpha = 0.25, colour = "steelblue") +
  stat_summary(aes(group = 1), fun = mean, geom = "line",
               colour = "red", linewidth = 1.2) +
  labs(title = "HbA1c Individual Trajectories (n = 100 patients; red = group mean)",
       x = "Time Point", y = "HbA1c (%)") +
  theme_gee()
```

A descending red mean line confirms HbA1c improves on average over time. The wide spread of individual blue lines demonstrates strong **between-patient heterogeneity** in baseline HbA1c, quantified later by the ICC. Lines that cross (non-parallel trajectories) suggest patients also differ in their rate of change, motivating the random slope extension in LMM.

### Mean Trajectory by Intervention Strength

```{r fig-trajectory, fig.cap="Figure 2. Mean HbA1c over time by intervention strength. Diverging lines indicate a Time × Intervention interaction: the effect of intervention on HbA1c differs across time points, motivating interaction models in both LMM and GEE."}
#| fig-width: 9
#| fig-height: 5

df |>
  group_by(time_step, intervention_strength) |>
  summarise(mean_hba1c = mean(hba1c), .groups = "drop") |>
  ggplot(aes(x = time_step, y = mean_hba1c,
             colour = intervention_strength, group = intervention_strength)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  scale_colour_manual(values = cb_intervention) +
  labs(title = "Mean HbA1c Over Time by Intervention Strength",
       x = "Time Point", y = "Mean HbA1c (%)",
       colour = "Intervention") +
  theme_gee()
```

If lines run **parallel**, there is no interaction, the intervention effect is constant across time points. Diverging lines (Intensive dropping fastest) signal a **time × intervention interaction**: the benefit of higher-intensity intervention accumulates over time, which will be formally tested in both the LMM and GEE interaction models.

### HbA1c Distribution by Group

```{r fig-violin, fig.cap="Figure 3. Distribution of HbA1c (%) by time point and intervention strength (n = 60,000 observations). Violin + box plots reveal distributional shape, spread, and temporal trend simultaneously. Colourblind-safe palette (Okabe & Ito, 2008)."}
df |>
  ggplot(aes(x = time_step, y = hba1c,
             fill = intervention_strength,
             colour = intervention_strength)) +
  geom_violin(alpha = 0.35, trim = TRUE, linewidth = 0.5) +
  geom_boxplot(alpha = 0.75, width = 0.22, outlier.size = 0.3,
               position = position_dodge(0.9)) +
  scale_fill_manual(values = cb_intervention, name = "Intervention Strength") +
  scale_colour_manual(values = cb_intervention, name = "Intervention Strength") +
  labs(
    title    = "HbA1c Distribution by Time Point and Intervention Strength",
    subtitle = "Violin + box plots | 20,000 patients, 3 time points",
    x = "Time Point", y = "HbA1c (%)"
  ) +
  theme_gee()
```

::: {.callout-tip}
## Figure 3 Interpretation
At T1\_Pre, all three groups show near-identical median HbA1c (~8.5–8.6%), confirming baseline group comparability. A progressive downward shift is visible across all groups from T1\_Pre to T3\_Post. By T3\_Post, the Intensive group (cobalt blue) has visibly lower median HbA1c than Standard, which in turn is lower than Weak, indicating a dose–response relationship that motivates the formal interaction analysis. The stable interquartile range across time points supports the homoscedasticity assumption for both LMM and GEE.
:::

### Correlation Matrix

```{r fig-corr, fig.cap="Figure 4. Pairwise Pearson correlations among all continuous variables. Colour intensity indicates strength (blue = positive, red = negative). Pairs with |r| > 0.8 would signal multicollinearity requiring variable removal."}
#| fig-width: 7
#| fig-height: 6

cont_vars <- df |>
  dplyr::select(where(is.double))

cor_matrix <- cor(as.matrix(cont_vars), use = "complete.obs")
corrplot::corrplot(cor_matrix, method = "color", type = "upper",
                   addCoef.col = "black", tl.col = "black", tl.srt = 45)
```

No pairwise correlation exceeds |r| = 0.8, confirming the absence of severe multicollinearity among the available continuous predictors.

### Table 1: Baseline Patient Characteristics

```{r table1}
df |>
  filter(time_step == "T1_Pre") |>
  dplyr::select(age, gender, bmi, hba1c,
                physical_activity_level, diet_score, sleep_hours,
                medication_adherence, intervention_strength,
                comorbidity_level) |>
  tbl_summary(
    statistic = list(
      all_continuous()  ~ "{mean} ({sd})",
      all_categorical() ~ "{n} ({p}%)"
    ),
    missing = "no",
    label = list(
      age                     ~ "Age (years)",
      gender                  ~ "Gender",
      hba1c                   ~ "HbA1c (%)",
      bmi                     ~ "BMI (kg/m²)",
      diet_score              ~ "Diet Score",
      sleep_hours             ~ "Sleep Hours",
      physical_activity_level ~ "Physical Activity Level",
      medication_adherence    ~ "Medication Adherence",
      intervention_strength   ~ "Intervention Strength",
      comorbidity_level       ~ "Comorbidity Level"
    )
  ) |>
  bold_labels() |>
  modify_caption("**Table 1. Baseline Characteristics at T1_Pre (n = 20,000)**")
```

Table 1 characterises the study cohort at baseline (T1\_Pre only). Filtering to baseline is essential, including all three time points would inflate n to 60,000, misrepresenting the actual cohort size of 20,000 patients. This table establishes the starting point before any intervention effects have occurred.

### Table 2: Longitudinal Descriptive Statistics by Time Point

```{r table2}
df |>
  dplyr::select(hba1c, time_step, intervention_strength,
                age, gender, bmi, physical_activity_level,
                diet_score, sleep_hours, medication_adherence,
                comorbidity_level) |>
  tbl_summary(
    by = time_step,
    statistic = list(
      all_continuous()  ~ "{mean} ({sd})",
      all_categorical() ~ "{n} ({p}%)"
    ),
    digits = list(
      all_continuous()  ~ 2,
      all_categorical() ~ c(0, 1)
    ),
    label = list(
      hba1c                   ~ "HbA1c (%)",
      intervention_strength   ~ "Intervention Strength",
      age                     ~ "Age (years)",
      gender                  ~ "Gender",
      bmi                     ~ "BMI (kg/m²)",
      physical_activity_level ~ "Physical Activity Level",
      diet_score              ~ "Diet Score",
      sleep_hours             ~ "Sleep Hours",
      medication_adherence    ~ "Medication Adherence",
      comorbidity_level       ~ "Comorbidity Level"
    ),
    missing = "no"
  ) |>
  add_overall() |>
  bold_labels() |>
  modify_header(label ~ "**Variable**") |>
  modify_caption("**Table 2. Descriptive Characteristics by Time Point (n = 60,000 observations)**")
```

::: {.callout-tip}
## Table 2 Interpretation
Mean HbA1c declined progressively from 8.71% at baseline to 8.00% post-intervention, suggesting a genuine temporal treatment effect. Patient-level fixed characteristics, gender, comorbidity, and age, remained constant across time points as expected. Time-varying behavioural variables showed favourable trends: physical activity (High category: 21.8% → 33.0%), diet score (5.20 → 5.67), and sleep hours (6.60 → 6.74). Mean BMI was 31.3 kg/m², consistent with an obese T2DM cohort.
:::

---

# Statistical Assumptions & Analytical Framework

Before fitting any model, two foundational assumptions must be evaluated: the distributional properties of the outcome, and the degree of within-patient clustering. These checks collectively determine which analytical method is required and whether ordinary regression would be sufficient.

## Normality of the Outcome

```{r fig-normality, fig.cap="Figure 5. Left: Q-Q plot of HbA1c values. Right: Frequency distribution histogram. A bell-shaped symmetric histogram and points close to the Q-Q diagonal indicate approximate normality."}
#| fig-width: 10
#| fig-height: 4

par(mfrow = c(1, 2))
qqnorm(df$hba1c, main = "Q-Q Plot: HbA1c")
qqline(df$hba1c, col = "red")
hist(df$hba1c, breaks = 50, main = "Distribution of HbA1c",
     xlab = "HbA1c (%)", col = "steelblue", border = "white")
par(mfrow = c(1, 1))
```

Both panels support approximate normality: the Q-Q plot shows points tightly aligned to the diagonal, and the histogram reveals a symmetric, bell-shaped distribution centred near the grand mean.

> **Note on sample size:** With n = 60,000 observations, the Shapiro-Wilk test is not appropriate, at large n it will always reject H₀ for trivially small, clinically irrelevant deviations from normality. Visual inspection is the appropriate approach. Both LMM and GEE are also robust to mild normality violations at large samples due to the Central Limit Theorem.

## Clustering: Intraclass Correlation Coefficient (ICC)

To determine whether the repeated-measures structure requires specialised modelling, an **intercept-only (null) LMM** is fitted. The ICC derived from this model quantifies the proportion of total HbA1c variance attributable to stable between-patient differences.

$$\text{ICC} = \frac{\sigma^2_{\text{between}}}{\sigma^2_{\text{between}} + \sigma^2_{\text{within}}}$$

```{r icc-null}
m_null <- lmer(hba1c ~ 1 + (1 | patient_id), data = df, REML = TRUE)
performance::icc(m_null)
```

**Result:** ICC = 0.813. This means **81.3%** of the total variance in HbA1c is attributable to stable, persistent differences between patients, only 18.7% represents within-patient change over time. This extremely high ICC confirms that ignoring the repeated-measures structure (i.e., running ordinary MLR on all 60,000 rows) would drastically underestimate standard errors and produce many spurious significant findings. Both LMM and GEE are therefore strongly justified.

## Choosing Between LMM and GEE: Complementary Frameworks

::: {.callout-note}
## LMM vs GEE: Two Answers to Two Different Questions

| Feature | LMM | GEE |
|---|---|---|
| **Estimand** | Subject-specific (individual-level) | Population-averaged (marginal) |
| **Research question** | "How does *this patient's* HbA1c change?" | "What is the expected HbA1c change for a *typical patient*?" |
| **Random effects** | Yes, random intercept ± slope | No, correlation handled via working matrix |
| **Interpretation** | Patient-level trajectory | Population-level policy estimate |
| **Suitable for prediction?** | Yes (individual patients) | No (population only) |
| **Correlation structure** | Implicitly via random effects | Explicitly via working matrix (QIC-selected) |
| **Public health relevance** | Understanding heterogeneity | Programme evaluation & policy |

In this analysis, both methods are applied. LMM provides subject-specific estimates and quantifies the heterogeneity in patient trajectories. GEE provides population-averaged estimates directly interpretable for national T2DM programme evaluation. The two models are complementary, not competing.
:::

## GEE-Specific Assumptions

::: {.callout-note}
## GEE Has Minimal Formal Assumptions
Unlike MLR or LMM, GEE does not require normality of the outcome or residuals, nor does it assume a distribution for random effects. The key assumptions for valid GEE inference are:

1. **Correct mean model specification**: the linear predictor correctly captures the relationship between predictors and HbA1c. *Verified via Pearson residual plots in Section 5.*
2. **Independence between clusters**: HbA1c measurements from different patients are independent. *Satisfied by study design.*
3. **Constant variance (Gaussian family)**: appropriate for continuous HbA1c. *Supported by stable residual spread in Section 5.*
4. **Large sample size**: the sandwich variance estimator is asymptotically valid. *Satisfied: n = 20,000 patients.*
5. **MCAR (if missing data present)**: GEE requires data to be Missing Completely At Random. *No missing values in this dataset.*

There is **no assumption** about the correctness of the working correlation structure; the sandwich estimator protects inference regardless of structure choice.
:::

---

# Linear Mixed Effects Model Analysis

The LMM extends standard linear regression by adding **random effects**, parameters that vary across higher-level grouping units (patients). This correctly partitions within-patient change from stable between-patient differences, as warranted by the ICC = 0.813 established in Section 3.

The general LMM formula fitted here is:

$$\text{HbA1c}_{ij} = \underbrace{(\beta_0 + u_{0i})}_{\text{patient-specific intercept}} + \beta_1 \cdot \text{time}_{ij} + \beta_2 \cdot \mathbf{X}_{ij} + \varepsilon_{ij}$$

Where:

- $\beta_0$ = population-average HbA1c intercept
- $u_{0i} \sim N(0, \sigma^2_{\text{between}})$ = patient-specific random intercept deviation
- $\beta_1, \beta_2$ = fixed effects for time and covariates
- $\varepsilon_{ij} \sim N(0, \sigma^2_{\text{within}})$ = within-patient residual

A **model-building strategy** is employed: four progressively richer models are fitted and compared using likelihood ratio tests and AIC. This approach isolates the contribution of each component and justifies the final model choice on statistical grounds.

## A: Model Building

### Null Model (Variance Decomposition)

The null model contains no predictors, only an overall intercept and a random intercept per patient. Its purpose is to decompose total HbA1c variance into between-patient and within-patient components, establishing the baseline ICC.

```{r lmm-null}
lmm_null <- lmer(hba1c ~ 1 + (1 | patient_id), data = df, REML = TRUE)
summary(lmm_null)
performance::icc(lmm_null)
```

### Time-Only Model (Unconditional Growth)

The time-only model adds `time_step` as the sole fixed-effect predictor. This addresses a foundational question: **does HbA1c change significantly over time at all?** The intercept estimates mean HbA1c at T1\_Pre; coefficients for T2\_Mid and T3\_Post represent the change from baseline averaged across all patients.

```{r lmm-time}
lmm_time <- lmer(hba1c ~ time_step + (1 | patient_id),
                 data = df, REML = TRUE)
summary(lmm_time)
```

**Result:** Both T2\_Mid (−0.40%) and T3\_Post (−0.71%) show highly significant declines from baseline (both p < 0.0001), confirming a meaningful time trend before accounting for any covariates.

### Full Fixed-Effects Model

This is the **main adjusted model**, it adds all available covariates alongside time, providing simultaneous estimates of each predictor's independent association with HbA1c after controlling for all others. `lme4::isSingular()` verifies the model has not been over-parameterised.

```{r lmm-full}
lmm_full <- lmer(
  hba1c ~ time_step + age + gender + bmi +
    physical_activity_level + diet_score + sleep_hours +
    medication_adherence + intervention_strength + comorbidity_level +
    (1 | patient_id),
  data = df, REML = TRUE
)

summary(lmm_full)
lme4::isSingular(lmm_full)   # check for over-parameterisation
```

**Result:** `isSingular() = FALSE`, the model converged cleanly. All fixed-effect predictors except gender (Male) and intervention (Standard) are statistically significant.

### Random Slope Extension

The random intercept model assumes all patients share the same average rate of change over time. The random slope model relaxes this: `(1 + time_num | patient_id)` allows both the intercept (starting level) and the slope (rate of change) to vary across patients.

```{r lmm-slope}
lmm_slope <- lmer(
  hba1c ~ time_step + age + gender + bmi +
    physical_activity_level + diet_score + sleep_hours +
    medication_adherence + intervention_strength + comorbidity_level +
    (1 + time_num | patient_id),
  data = df, REML = TRUE
)

lme4::isSingular(lmm_slope)
```

**Result:** `isSingular() = FALSE`, the random slope model is supported by the data, confirming that patients differ not only in their baseline HbA1c but also in their rate of change over time. However, because the primary clinical question concerns group-level trajectories, the interaction model below is the more relevant final model.

### Interaction Model: Time × Intervention

This model directly addresses the **primary research question** by testing whether intervention strength groups follow different HbA1c trajectories over time. The `*` operator creates both main effects and all pairwise interaction terms:

`time_step * intervention_strength` expands to:
`time_step + intervention_strength + time_step:intervention_strength`

::: {.callout-important}
## Critical Interpretation Rule
When a statistically significant interaction is present, **main effects must not be interpreted in isolation**. The main effect of `time_step` describes only the Weak group (reference); the main effect of `intervention_strength` describes only T1\_Pre (reference). Always use `emmeans()` to obtain properly marginalised cell-level estimates when an interaction is present.
:::

```{r lmm-interaction}
lmm_int <- lmer(
  hba1c ~ time_step * intervention_strength +
    age + gender + bmi + physical_activity_level + diet_score +
    sleep_hours + medication_adherence + comorbidity_level +
    (1 | patient_id),
  data = df, REML = TRUE
)

summary(lmm_int)
```

**Result:** All four interaction terms are highly significant (all p < 0.0001), confirming that intervention groups diverge in their HbA1c trajectories over time. The Intensive group shows the steepest decline.

## A: Model Comparison

Formal model comparison determines which of the four models best fits the data. Two complementary approaches are used:

1. **Likelihood Ratio Test (LRT)** via `anova()`: valid for nested models.
2. **AIC (Akaike Information Criterion)**: valid for both nested and non-nested models; lower AIC = better balance of fit and parsimony; a difference > 10 is considered strong evidence.

> **Why refit with REML = FALSE?** REML cannot validly compare models with different fixed-effect structures. ML must be used for fixed-effects comparisons; REML is used only for final parameter estimates.

```{r lmm-comparison}
lmm_null_ml  <- update(lmm_null,  REML = FALSE)
lmm_time_ml  <- update(lmm_time,  REML = FALSE)
lmm_full_ml  <- update(lmm_full,  REML = FALSE)
lmm_int_ml   <- update(lmm_int,   REML = FALSE)

anova(lmm_null_ml, lmm_time_ml, lmm_full_ml)
AIC(lmm_null_ml, lmm_time_ml, lmm_full_ml, lmm_int_ml)
```

| Model | AIC | Improvement |
|---|---|---|
| Null | 154,667 | Baseline, no predictors |
| Time-only | 134,788 | −19,879; time highly significant |
| Full covariates | 101,920 | −32,868; covariates essential |
| Interaction | **101,119** | −801; interaction significant |

The interaction model achieves the lowest AIC and is confirmed as the best-fitting model. Each step's improvement is statistically significant (all p < 2.2×10⁻¹⁶ by LRT).

## A: Model Assessment

### Variance Explained: Marginal and Conditional R²

LMMs have two R² values (Nakagawa & Schielzeth, 2013):

- **R²m (marginal)** = proportion of total variance explained by fixed effects only
- **R²c (conditional)** = proportion explained by fixed + random effects combined

A large gap between R²m and R²c confirms the random intercept is absorbing substantial between-patient variance, validating the LMM approach.

```{r lmm-r2}
MuMIn::r.squaredGLMM(lmm_full)
performance::r2(lmm_full)
performance::icc(lmm_full)
```

**Results:** R²m = 0.368, R²c = 0.974. Fixed effects alone explain 36.8% of variance; once patient-level random intercepts are included, 97.4% is explained. The gap (~60%) represents stable between-patient differences that the measured covariates cannot fully capture, confirming LMM is strongly justified.

### Residual Diagnostics

```{r lmm-residuals, fig.cap="Figure 6. LMM residual diagnostics. Left: residuals vs fitted; random scatter around zero indicates no systematic misspecification. Right: Q-Q plot of residuals; approximate normality of residuals confirms model adequacy."}
#| fig-width: 6
#| fig-height: 5

plot(lmm_full)                                    # residuals vs fitted
qqnorm(resid(lmm_full)); qqline(resid(lmm_full))  # normality of residuals
```

Both panels confirm adequate model specification: no systematic pattern in the residual plot and points closely aligned to the Q-Q diagonal.

## A: Results: Estimated Marginal Means & Contrasts

### Within-Group Time Contrasts

When a significant interaction is present, `emmeans()` calculates model-predicted mean HbA1c at each cell of the time × intervention grid, **marginalised over** all other covariates. `contrast()` then performs all pairwise comparisons within each intervention group across time, with Bonferroni correction.

```{r lmm-emmeans}
emm_int <- emmeans::emmeans(lmm_int, ~ time_step | intervention_strength)
emm_int

emmeans::contrast(emm_int, method = "pairwise", adjust = "bonferroni")
```

```{r fig-lmm-emm, fig.cap="Figure 7. Estimated marginal mean HbA1c trajectories by intervention strength from the LMM interaction model (95% CI shaded). Non-parallel lines confirm a significant Time × Intervention interaction."}
#| fig-width: 8
#| fig-height: 5

emmeans::emmip(lmm_int, intervention_strength ~ time_step, CIs = TRUE) +
  scale_colour_manual(values = c("Weak"      = "#e74c3c",
                                 "Standard"  = "#f39c12",
                                 "Intensive" = "#27ae60")) +
  labs(
    title    = "LMM: HbA1c Trajectories by Intervention Strength",
    subtitle = "Estimated marginal means (95% CI shaded) - adjusted for all covariates",
    x        = "Time Point",
    y        = "HbA1c (%) - Estimated Marginal Mean",
    colour   = "Intervention Strength"
  ) +
  theme_bw(base_size = 12)
```

### Between-Group Contrasts at Endpoint (T3_Post)

This block answers the complementary question: **At the study endpoint, do the three intervention groups differ significantly from each other?** Bonferroni correction is applied across the three pairwise contrasts.

```{r lmm-t3-contrasts}
emm_t3 <- emmeans::emmeans(lmm_int, ~ intervention_strength | time_step,
                            at = list(time_step = "T3_Post"))
emm_t3

emmeans::contrast(emm_t3, method = "pairwise", adjust = "bonferroni")
```

At T3\_Post, the estimated marginal HbA1c means are:

- **Weak**: 8.238% (95% CI: 8.197–8.279)
- **Standard**: 8.158% (95% CI: 8.130–8.186)
- **Intensive**: 8.036% (95% CI: 8.001–8.071)

All three pairwise comparisons are statistically significant after Bonferroni correction. Intensive intervention achieves HbA1c 0.202% lower than Weak (p < 0.0001) and 0.122% lower than Standard (p < 0.0001). Even Standard confers a significant advantage over Weak (p = 0.0039), confirming a dose–response gradient across all three tiers at the study endpoint.

## P: Final Results: Model Table & Forest Plot

### Final LMM Model Table

The table below is the primary results table for the LMM analysis, reporting fixed-effect estimates from the best-fitting model (`lmm_int`) with 95% confidence intervals, p-values, random-effect variance components, ICC, R²m, R²c, and AIC.

```{r lmm-final-table}
sjPlot::tab_model(
  lmm_int,
  show.re.var = TRUE,
  show.icc    = TRUE,
  show.r2     = TRUE,
  show.aic    = TRUE,
  pred.labels = c(
    "(Intercept)",
    "Time: T2_Mid",
    "Time: T3_Post",
    "Intervention: Standard",
    "Intervention: Intensive",
    "Age",
    "Gender: Male",
    "BMI",
    "Physical Activity: Moderate",
    "Physical Activity: High",
    "Diet Score",
    "Sleep Hours",
    "Medication Adherence: Moderate",
    "Medication Adherence: Good",
    "Comorbidity: Medium",
    "Comorbidity: High",
    "T2_Mid × Standard",
    "T3_Post × Standard",
    "T2_Mid × Intensive",
    "T3_Post × Intensive"
  ),
  title = "Table 3: LMM Interaction Model - Predictors of HbA1c (n = 60,000 observations; 20,000 patients)"
)
```

### Forest Plot: Fixed Effects

```{r fig-forest, fig.cap="Figure 8. Forest plot of fixed-effect estimates from lmm_full (95% CI). Points left of the dashed zero line (red) indicate HbA1c-lowering effects; points to the right indicate HbA1c-raising effects. Intervals crossing zero indicate non-significant predictors."}
#| fig-width: 8
#| fig-height: 6

sjPlot::plot_model(
  lmm_full,
  type         = "est",
  show.values  = TRUE,
  show.p       = TRUE,
  value.offset = 0.35,
  title = "Fixed-Effect Estimates from LMM (lmm_full)\nReference: T1_Pre, Female, Low Activity, Poor Adherence, Weak Intervention, Low Comorbidity"
) +
  theme_bw(base_size = 11) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey40")
```

### Model Fit Summary

```{r lmm-fit-summary}
r2_lmm  <- MuMIn::r.squaredGLMM(lmm_full)
lmm_r2m <- r2_lmm[1, "R2m"]
lmm_r2c <- r2_lmm[1, "R2c"]
icc_val <- as.numeric(performance::icc(lmm_null)$ICC_adjusted)
```

| Metric | Value | Interpretation |
|---|---|---|
| ICC (null model) | 0.813 | 81.3% of HbA1c variance is between patients: LMM required |
| R²m (marginal) | 0.368 | Fixed effects explain 36.8% of total variance |
| R²c (conditional) | 0.974 | Fixed + random effects explain 97.4% of total variance |
| AIC (lmm_int) | 101,119 | Best-fitting model across all four candidates |

## LMM Conclusions

This analysis examined longitudinal HbA1c trajectories in 20,000 patients across three time points using LMM. The null model ICC of **0.813** confirmed that 81.3% of total HbA1c variance is attributable to stable between-patient differences, validating the LMM framework.

All four time × intervention interaction terms were statistically significant (all p < 0.0001). The Intensive intervention produced the steepest HbA1c decline (mean 8.04% at T3\_Post vs 8.24% for Weak). The key independent modifiable predictors were: physical activity High (−0.52%***), sleep hours (−0.41%***), medication adherence Good (−0.36%***), and diet score (−0.32%***). BMI (+0.09%***) was the most important detrimental factor. At T3\_Post, all three intervention groups differed significantly from each other after Bonferroni correction.

---

# Generalised Estimating Equations Analysis

Whereas LMM estimates subject-specific effects, GEE estimates **population-averaged (marginal) effects**, the expected HbA1c change for a typical patient drawn from the T2DM population. These coefficients directly answer policy-relevant questions: "If all Malaysian T2DM patients received Intensive versus Weak intervention, what is the expected population-level HbA1c difference?" The GEE estimand is therefore the appropriate primary method for public health programme evaluation.

## GEE Theoretical Framework

### The Marginal (Population-Averaged) Model

GEE estimates the **population-averaged** expectation of HbA1c. For a continuous outcome with an identity link:

$$
\begin{aligned}
E(Y_{ij} \mid \mathbf{X}_{ij}) = \; &\beta_0
+ \beta_1 \, \text{T2 Mid}_{ij} + \beta_2 \, \text{T3 Post}_{ij} \\
&+ \beta_3 \, \text{Standard}_{ij} + \beta_4 \, \text{Intensive}_{ij} \\
&+ \beta_5 \, \text{Age}_{ij} + \beta_6 \, \text{Male}_{ij} + \beta_7 \, \text{BMI}_{ij} \\
&+ \beta_8 \, \text{PA Moderate}_{ij} + \beta_9 \, \text{PA High}_{ij} \\
&+ \beta_{10} \, \text{Diet Score}_{ij} + \beta_{11} \, \text{Sleep Hours}_{ij} \\
&+ \beta_{12} \, \text{MedAdh Moderate}_{ij} + \beta_{13} \, \text{MedAdh Good}_{ij} \\
&+ \beta_{14} \, \text{Comorbidity Medium}_{ij} + \beta_{15} \, \text{Comorbidity High}_{ij}
\end{aligned}
$$

Each $\beta$ coefficient represents the **expected change in HbA1c (%)** across the entire T2DM population for a one-unit change in the predictor, holding all other variables constant.

### Working Correlation Structures

Four structures were fitted and compared using QIC. The working correlation matrix $\mathbf{R}_i(\alpha)$ defines the assumed within-patient correlation pattern:

| Structure | `corstr =` | Assumption | Role |
|---|---|---|---|
| Independence | `"independence"` | All off-diagonal = 0 | Baseline reference |
| Exchangeable | `"exchangeable"` | All pairs = $\alpha$ | Primary candidate |
| AR(1) | `"ar1"` | Correlation decays: $\alpha^{|j-k|}$ | Sensitivity check |
| Unstructured | `"unstructured"` | Each pair has its own $\alpha_{jk}$ | Flexible benchmark |

### QIC Model Selection Criterion

The **Quasi-likelihood under Independence Criterion** (Pan, 2001) is the GEE analogue of AIC. Lower QIC indicates a better-fitting correlation structure.

$$\text{QIC} = -2\,\hat{Q}(\hat{\boldsymbol{\beta}}) + 2\,\text{tr}(\hat{\boldsymbol{\Omega}}_I \hat{\mathbf{V}}_R)$$

The first term ($-2\hat{Q}$) measures goodness of fit; the second term ($2\text{tr}(\hat{\boldsymbol{\Omega}}_I \hat{\mathbf{V}}_R)$) penalises for complexity. A more complex structure (e.g., Unstructured with 3 free $\alpha$ parameters) must improve fit substantially to overcome this penalty.

### Robustness: The Sandwich Variance Estimator

A critical property of GEE is that **valid estimates and correct standard errors are obtained even if the working correlation structure is misspecified**, provided the robust sandwich variance estimator is used (Liang & Zeger, 1986). This means β estimates remain consistent and standard errors are corrected using observed residuals, not model assumptions. QIC guides the best-fitting structure, but inference is protected even if the chosen structure is imperfect.

## A: Univariate GEE Screening

Univariate GEE with Exchangeable working correlation screens each predictor individually. Variables with p < 0.25 are retained for multivariable modelling.

```{r univariate-gee}
library(purrr)

predictors <- c(
  "time_step", "intervention_strength",
  "age", "gender", "bmi",
  "physical_activity_level", "diet_score",
  "sleep_hours", "medication_adherence",
  "comorbidity_level"
)

uni_list <- map(predictors, function(pred) {
  fml <- as.formula(paste("hba1c ~", pred))
  geeglm(
    fml,
    id     = patient_id,
    data   = df,
    family = gaussian(link = "identity"),
    corstr = "exchangeable"
  )
})

names(uni_list) <- predictors

uni_results <- map_dfr(names(uni_list), function(pred) {
  broom::tidy(uni_list[[pred]], conf.int = TRUE) |>
    filter(term != "(Intercept)") |>
    mutate(predictor = pred)
}) |>
  mutate(
    Predictor = case_match(term,
      "time_stepT2_Mid"                      ~ "Time: T2_Mid vs T1_Pre",
      "time_stepT3_Post"                     ~ "Time: T3_Post vs T1_Pre",
      "intervention_strengthStandard"        ~ "Intervention: Standard vs Weak",
      "intervention_strengthIntensive"       ~ "Intervention: Intensive vs Weak",
      "age"                                  ~ "Age (years)",
      "genderMale"                           ~ "Gender: Male vs Female",
      "bmi"                                  ~ "BMI (kg/m²)",
      "physical_activity_levelModerate"      ~ "Physical Activity: Moderate vs Low",
      "physical_activity_levelHigh"          ~ "Physical Activity: High vs Low",
      "diet_score"                           ~ "Diet Score",
      "sleep_hours"                          ~ "Sleep Hours",
      "medication_adherenceModerate"         ~ "Medication Adherence: Moderate vs Poor",
      "medication_adherenceGood"             ~ "Medication Adherence: Good vs Poor",
      "comorbidity_levelMedium"              ~ "Comorbidity: Medium vs Low",
      "comorbidity_levelHigh"                ~ "Comorbidity: High vs Low",
      .default                               = term
    ),
    `Crude β (95% CI)` = sprintf("%.3f (%.3f, %.3f)",
                                  estimate, conf.low, conf.high),
    `p-value` = case_when(
      p.value < 0.001 ~ "< 0.001***",
      p.value < 0.01  ~ sprintf("%.3f**", p.value),
      p.value < 0.05  ~ sprintf("%.3f*",  p.value),
      TRUE            ~ sprintf("%.3f",    p.value)
    ),
    Significant = ifelse(p.value < 0.05, "✓", "")
  ) |>
  select(Predictor, `Crude β (95% CI)`,
         `Robust SE` = std.error, `p-value`, Significant)
```

```{r univariate-table}
knitr::kable(
  uni_results,
  digits  = 3,
  caption = "Table 4. Univariate GEE Associations with HbA1c (%) (Exchangeable Working Correlation)"
) |>
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width        = FALSE
  ) |>
  pack_rows("Time Effects",          1, 2)  |>
  pack_rows("Intervention Effects",  3, 4)  |>
  pack_rows("Socio-demographic",     5, 7)  |>
  pack_rows("Lifestyle Factors",     8, 11) |>
  pack_rows("Clinical Factors",     12, 15) |>
  column_spec(3, bold = TRUE, color = "#009E73") |>
  footnote(general = c(
    "Crude β = unadjusted population-averaged change in HbA1c (%) per unit change in predictor.",
    "Each row is from a separate GEE model with only that predictor (simple GEE).",
    "Exchangeable working correlation; robust sandwich variance estimator.",
    "Reference categories: time_step = T1_Pre; intervention = Weak; gender = Female;",
    "  physical_activity = Low; medication_adherence = Poor; comorbidity = Low.",
    "✓ = significant at p < 0.05.",
    "* p < 0.05   ** p < 0.01   *** p < 0.001"
  ))
```

```{r fig-univariate-plot, fig.height=6.5, fig.cap="Figure 9. Univariate GEE crude coefficient plot. Filled circles = significant (p < 0.05); open circles = non-significant. Gender (Male) is the only non-significant predictor; all others proceed to multivariable adjustment."}
uni_plot <- map_dfr(names(uni_list), function(pred) {
  broom::tidy(uni_list[[pred]], conf.int = TRUE) |>
    filter(term != "(Intercept)") |>
    mutate(predictor = pred)
}) |>
  mutate(
    label = case_match(term,
      "time_stepT2_Mid"                      ~ "Time: T2_Mid",
      "time_stepT3_Post"                     ~ "Time: T3_Post",
      "intervention_strengthStandard"        ~ "Intervention: Standard",
      "intervention_strengthIntensive"       ~ "Intervention: Intensive",
      "age"                                  ~ "Age",
      "genderMale"                           ~ "Gender: Male",
      "bmi"                                  ~ "BMI",
      "physical_activity_levelModerate"      ~ "PA: Moderate",
      "physical_activity_levelHigh"          ~ "PA: High",
      "diet_score"                           ~ "Diet Score",
      "sleep_hours"                          ~ "Sleep Hours",
      "medication_adherenceModerate"         ~ "Med Adherence: Moderate",
      "medication_adherenceGood"             ~ "Med Adherence: Good",
      "comorbidity_levelMedium"              ~ "Comorbidity: Medium",
      "comorbidity_levelHigh"                ~ "Comorbidity: High",
      .default                               = term
    ),
    group = case_when(
      str_detect(term, "time_step")              ~ "Time",
      str_detect(term, "intervention_strength")  ~ "Intervention",
      term %in% c("age", "genderMale", "bmi")    ~ "Socio-demographic",
      str_detect(term, "physical_activity|diet|sleep") ~ "Lifestyle",
      str_detect(term, "medication|comorbidity") ~ "Clinical"
    ),
    significant = p.value < 0.05
  )

group_colours <- c(
  "Time"              = "#0072B2",
  "Intervention"      = "#009E73",
  "Socio-demographic" = "#E69F00",
  "Lifestyle"         = "#56B4E9",
  "Clinical"          = "#D55E00"
)

ggplot(uni_plot,
       aes(x = estimate, y = reorder(label, estimate),
           colour = group, shape = significant)) +
  geom_vline(xintercept = 0, linetype = "dashed",
             colour = "#D55E00", linewidth = 0.8) +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high),
                 height = 0.3, linewidth = 0.7) +
  geom_point(size = 3.5) +
  scale_colour_manual(values = group_colours, name = "Predictor Group") +
  scale_shape_manual(values = c("TRUE" = 16, "FALSE" = 1),
                     name   = "Significant (p < 0.05)") +
  labs(
    title    = "Univariate GEE Crude Coefficient Plot",
    subtitle = "Simple GEE (Exchangeable) | Crude β with 95% CI",
    x        = "Crude β (Change in HbA1c, %)",
    y        = NULL
  ) +
  theme_gee(base_size = 11) +
  theme(legend.position = "right")
```

::: {.callout-tip}
## Figure 9 Interpretation
Lifestyle factors (sky blue) dominate the crude associations: Medication Adherence Good, Physical Activity High, diet score, and sleep hours each show large, significant crude negative effects. These will be important confounders to adjust for in the multivariable model. Gender (Male) is the only non-significant predictor (CI crosses zero, open circle), confirmed also in the multivariable model. Despite non-significance, Gender is retained as an a priori biological confounder of glycaemic control.
:::

## A: Multivariable GEE: Four Correlation Structures

Four multivariable GEE models are fitted simultaneously, each containing all predictors but with a different working correlation structure. All predictors identified in univariate screening are retained, including Gender, given its established role as a biological confounder.

```{r gee-models}
gee_ind <- geeglm(
  hba1c ~ time_step + intervention_strength +
           age + gender + bmi +
           physical_activity_level + diet_score +
           sleep_hours + medication_adherence +
           comorbidity_level,
  id     = patient_id,
  data   = df,
  family = gaussian(link = "identity"),
  corstr = "independence"
)

gee_exch <- geeglm(
  hba1c ~ time_step + intervention_strength +
           age + gender + bmi +
           physical_activity_level + diet_score +
           sleep_hours + medication_adherence +
           comorbidity_level,
  id     = patient_id,
  data   = df,
  family = gaussian(link = "identity"),
  corstr = "exchangeable"
)

gee_ar1 <- geeglm(
  hba1c ~ time_step + intervention_strength +
           age + gender + bmi +
           physical_activity_level + diet_score +
           sleep_hours + medication_adherence +
           comorbidity_level,
  id     = patient_id,
  data   = df,
  family = gaussian(link = "identity"),
  corstr = "ar1"
)

gee_unstr <- geeglm(
  hba1c ~ time_step + intervention_strength +
           age + gender + bmi +
           physical_activity_level + diet_score +
           sleep_hours + medication_adherence +
           comorbidity_level,
  id     = patient_id,
  data   = df,
  family = gaussian(link = "identity"),
  corstr = "unstructured"
)

cat("All four GEE models fitted successfully.\n")
cat(sprintf("Unstructured α parameters: α12=%.4f  α13=%.4f  α23=%.4f\n",
            gee_unstr$geese$alpha[1],
            gee_unstr$geese$alpha[2],
            gee_unstr$geese$alpha[3]))
```

The raw summaries for each structure are presented below. These allow comparison of point estimates under different correlation assumptions before formal QIC-based selection in the next section. Inference should not be drawn from any single summary here; the QIC-selected model is the basis for final inference.

### Independence (Baseline Reference)

```{r gee-ind-summary}
summary(gee_ind)
```

::: {.callout-note collapse="true"}
## Independence Model Note
Independence GEE is equivalent to OLS regression with robust sandwich SE. It ignores within-patient correlation entirely, presented here as the formal baseline against which correlated structures are compared.
:::

### Exchangeable

```{r gee-exch-summary}
summary(gee_exch)
```

### AR(1): Autoregressive Order 1

```{r gee-ar1-summary}
summary(gee_ar1)
```

### Unstructured (Flexible Reference)

```{r gee-unstr-summary}
summary(gee_unstr)
cat("\nUnstructured α parameters (α₁₂, α₁₃, α₂₃):\n")
print(gee_unstr$geese$alpha)
```

::: {.callout-note collapse="true"}
## Reading These Summaries
Differences in point estimates and p-values across the four summaries reflect each structure's correlation assumption. The formal selection criterion (QIC) in the next section determines which structure produces the most parsimonious fit.
:::

## A: QIC-Based Model Selection

```{r qic-comparison}
qic_ind   <- QIC(gee_ind)
qic_exch  <- QIC(gee_exch)
qic_ar1   <- QIC(gee_ar1)
qic_unstr <- QIC(gee_unstr)

qic_vals <- c(
  qic_ind["QIC"], qic_exch["QIC"],
  qic_ar1["QIC"], qic_unstr["QIC"]
)
best_row <- which.min(qic_vals)

qic_table <- tibble(
  `Correlation Structure` = c("Independence", "Exchangeable", "AR(1)", "Unstructured"),
  `Free α Parameters`     = c("0", "1", "1", "3 (α₁₂, α₁₃, α₂₃)"),
  QIC  = c(qic_ind["QIC"],  qic_exch["QIC"],  qic_ar1["QIC"],  qic_unstr["QIC"]),
  QICu = c(qic_ind["QICu"], qic_exch["QICu"], qic_ar1["QICu"], qic_unstr["QICu"]),
  Selected = ifelse(qic_vals == min(qic_vals), "✓ Best", "")
)

cat(sprintf("QIC ranking - Independence: %.2f | Exchangeable: %.2f | AR(1): %.2f | Unstructured: %.2f\n",
            qic_vals[1], qic_vals[2], qic_vals[3], qic_vals[4]))
cat(sprintf("Best structure: %s (row %d)\n",
            qic_table$`Correlation Structure`[best_row], best_row))

knitr::kable(
  qic_table,
  digits  = 2,
  caption = "Table 5. QIC Comparison: Four GEE Working Correlation Structures (lower QIC = better fit)"
) |>
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = TRUE
  ) |>
  row_spec(best_row, bold = TRUE, color = "#0b2d5e", background = "#eaf3fb") |>
  footnote(general = c(
    "QIC = Quasi-likelihood under Independence Criterion (Pan, 2001).",
    "QICu = QIC under independence model (used for mean model selection).",
    "Lower QIC indicates a better-fitting working correlation structure.",
    "✓ Best = structure used for final inference."
  ))
```

::: {.callout-important}
## Table 5 Interpretation
The Exchangeable structure achieved the lowest QIC (44,871.88), indicating the best balance of fit and parsimony, and is selected as the primary model. Independence performed worst (QIC = 60,075.23), confirming that ignoring within-patient correlation is inappropriate. The Unstructured α parameters (α₁₂=0.940, α₁₃=0.909, α₂₃=0.977) showed a narrow range of 0.068, empirically confirming that the equal-correlation constraint of Exchangeable is justified. **The `gee_exch` model is the QIC-preferred model for all final GEE inference.**
:::

## A: Estimated Working Correlation Parameters

```{r gee-alpha}
cat(sprintf("Exchangeable α         : %.4f\n", gee_exch$geese$alpha))
cat(sprintf("AR(1) α                : %.4f\n", gee_ar1$geese$alpha))
cat(sprintf("Unstructured α12 (T1↔T2): %.4f\n", gee_unstr$geese$alpha[1]))
cat(sprintf("Unstructured α13 (T1↔T3): %.4f\n", gee_unstr$geese$alpha[2]))
cat(sprintf("Unstructured α23 (T2↔T3): %.4f\n", gee_unstr$geese$alpha[3]))
```

::: {.callout-tip}
## Working Correlation Interpretation
The Exchangeable α = **0.9590** and AR(1) α = **0.9583** are nearly identical, indicating that HbA1c correlations between any two time points from the same patient are approximately equal, consistent with HbA1c being a highly stable physiological marker. The slight temporal gradient in the Unstructured parameters (α₁₃ < α₁₂ < α₂₃, range = 0.068) suggests very minor decay over longer time gaps, but this does not overturn the Exchangeable QIC advantage. Critically, these extreme within-patient correlations (α ≈ 0.96) confirm that using MLR on these data would severely underestimate standard errors and inflate Type I error rates.
:::

## A: Interaction Analysis: Time × Intervention

When the effect of intervention on HbA1c differs by time point, a Time × Intervention interaction is required. Main effects alone cannot capture this divergence. The AR(1) structure was specified for the interaction model to better reflect the temporal nature of the hypothesis; results are compared across structures as sensitivity analysis.

### Interaction Model Fit

```{r gee-interaction}
gee_int <- geeglm(
  hba1c ~ time_step * intervention_strength +
           age + gender + bmi +
           physical_activity_level + diet_score +
           sleep_hours + medication_adherence +
           comorbidity_level,
  id     = patient_id,
  data   = df,
  family = gaussian(link = "identity"),
  corstr = "ar1"
)

cat("=== GEE Interaction Model: Time × Intervention ===\n")
summary(gee_int)
```

### Estimated Marginal Means

```{r gee-emmeans}
emm_gee <- emmeans(gee_int,
                   ~ time_step | intervention_strength,
                   level = 0.95)
print(emm_gee)

cat("\n=== Pairwise Time-Point Contrasts (Bonferroni-corrected) ===\n")
contrast(emm_gee, method = "pairwise", adjust = "bonferroni")
```

### EMM Summary Table

```{r gee-emm-table}
emm_df <- as.data.frame(confint(emm_gee, level = 0.95))

ci_lo <- grep("lower|lower\\.CL|LCL", names(emm_df), value = TRUE, ignore.case = TRUE)[1]
ci_hi <- grep("upper|upper\\.CL|UCL", names(emm_df), value = TRUE, ignore.case = TRUE)[1]

if (is.na(ci_lo)) {
  emm_df$ci_lo <- emm_df$emmean - 1.96 * emm_df$SE; ci_lo <- "ci_lo"
  emm_df$ci_hi <- emm_df$emmean + 1.96 * emm_df$SE; ci_hi <- "ci_hi"
}

emm_display <- emm_df |>
  mutate(
    `EMM (95% CI)` = sprintf("%.3f (%.3f, %.3f)",
                              emmean, .data[[ci_lo]], .data[[ci_hi]])
  ) |>
  select(`Time Point`   = time_step,
         `Intervention` = intervention_strength,
         `EMM (95% CI)`,
         SE)

knitr::kable(
  emm_display,
  digits  = 3,
  caption = "Table 6. GEE Estimated Marginal Means: HbA1c (%) by Time × Intervention Strength"
) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE) |>
  footnote(general = c(
    "EMM = population-averaged HbA1c (%) at each time–intervention combination.",
    "Adjusted for age, gender, BMI, physical activity, diet score, sleep hours,",
    "  medication adherence, and comorbidity level.",
    "AR(1) working correlation; robust sandwich standard errors."
  ))
```

::: {.callout-tip}
## Table 6 Interpretation
All three groups started at nearly identical HbA1c at T1\_Pre (~8.60–8.61%). By T3\_Post, the Intensive group achieved the lowest EMM at **7.96%**, followed by Standard (**8.14%**) and Weak (**8.26%**). The Intensive-vs-Weak gap widened from 0.01 pp at T1\_Pre to **0.30 pp** at T3\_Post, a 30-fold divergence demonstrating a clear dose–response relationship. All within-group pairwise time contrasts were significant after Bonferroni correction (p < 0.0001).
:::

### Interaction Trajectory Plot

```{r fig-gee-emm, fig.height=5.5, fig.cap="Figure 10. GEE estimated marginal mean HbA1c trajectories by intervention strength (AR(1) interaction model; 95% CI shaded). Non-parallel lines confirm significant Time × Intervention interaction. The Intensive group (cobalt blue) achieves the lowest HbA1c at T3_Post."}
ggplot(emm_df,
       aes(x      = time_step,
           y      = emmean,
           group  = intervention_strength,
           colour = intervention_strength,
           fill   = intervention_strength)) +
  geom_ribbon(aes(ymin = .data[[ci_lo]], ymax = .data[[ci_hi]]),
              alpha = 0.18, colour = NA) +
  geom_line(linewidth = 1.3) +
  geom_point(size = 4) +
  scale_colour_manual(values = cb_intervention, name = "Intervention Strength") +
  scale_fill_manual(values   = cb_intervention, name = "Intervention Strength") +
  scale_y_continuous(labels = scales::number_format(accuracy = 0.01)) +
  labs(
    title    = "GEE Estimated Marginal Mean HbA1c Trajectories",
    subtitle = "Time × Intervention interaction model (AR1 GEE) | 95% CI shaded",
    x        = "Time Point",
    y        = "Estimated Mean HbA1c (%)"
  ) +
  theme_gee()
```

## A: Model Diagnostics

### Pearson Residual Analysis

Traditional influential diagnostics (Cook's distance, leverage) are derived for independent observations and do not apply directly to GEE. With n = 20,000 clusters, no single patient can meaningfully shift population-averaged estimates. Model adequacy is assessed via Pearson residual plots.

```{r fig-gee-diagnostics, fig.cap="Figure 11. GEE diagnostic plots (Exchangeable structure). Left: Pearson residuals vs fitted; random scatter around zero confirms adequate mean model specification. Right: Q-Q plot of Pearson residuals; approximate normality is expected with n = 60,000."}
gee_resid  <- residuals(gee_exch, type = "pearson")
gee_fitted <- fitted(gee_exch)

par(mfrow = c(1, 2),
    mar   = c(4.5, 4.5, 3, 1.5),
    family = "sans")

plot(gee_fitted, gee_resid,
     main = "Pearson Residuals vs Fitted",
     xlab = "Fitted values (HbA1c, %)",
     ylab = "Pearson Residuals",
     col  = adjustcolor("#0072B2", 0.22),
     pch  = 20, cex = 0.45)
abline(h = 0, col = "#D55E00", lwd = 1.8, lty = 2)
lines(lowess(gee_fitted, gee_resid), col = "#D55E00", lwd = 2)
mtext("Loess smooth in vermillion | Ideal: flat around y = 0",
      side = 1, line = 4, cex = 0.72, col = "#6b7a99")

qqnorm(gee_resid,
       main = "Q-Q Plot: Pearson Residuals",
       xlab = "Theoretical Quantiles",
       ylab = "Sample Quantiles",
       col  = "#0072B2", pch = 20, cex = 0.45)
qqline(gee_resid, col = "#D55E00", lwd = 2)
mtext("Ideal: points on diagonal | Tails acceptable at large n",
      side = 1, line = 4, cex = 0.72, col = "#6b7a99")

par(mfrow = c(1, 1))
```

::: {.callout-note}
## Figure 11 Interpretation
**Left panel:** Pearson residuals scatter randomly around the zero reference line with no systematic curvature or funnel pattern. The Loess smooth runs near-horizontally confirming: (i) the mean model is correctly specified; (ii) no evidence of severe heteroscedasticity; and (iii) the Gaussian family with identity link is appropriate. **Right panel:** Sample quantiles closely follow the normal diagonal across the central range; minor tail deviations are expected at n = 60,000 and are of no practical concern given the asymptotic validity of the sandwich estimator.
:::

### Observed Mean Trajectory Verification

```{r fig-observed-trajectory, fig.cap="Figure 12. Observed mean HbA1c trajectories (±1 SD) by intervention group. This empirically validates that GEE model-estimated EMMs (Figure 10) faithfully reflect raw data patterns."}
df |>
  group_by(time_step, intervention_strength) |>
  summarise(
    mean_hba1c = mean(hba1c),
    sd_hba1c   = sd(hba1c),
    .groups    = "drop"
  ) |>
  ggplot(aes(x      = time_step,
             y      = mean_hba1c,
             group  = intervention_strength,
             colour = intervention_strength)) +
  geom_errorbar(aes(ymin = mean_hba1c - sd_hba1c,
                    ymax = mean_hba1c + sd_hba1c),
                width = 0.15, linewidth = 0.6,
                position = position_dodge(0.15)) +
  geom_line(linewidth = 1.1, position = position_dodge(0.15)) +
  geom_point(size = 3.5, position = position_dodge(0.15)) +
  scale_colour_manual(values = cb_intervention, name = "Intervention Strength") +
  labs(
    title    = "Observed Mean HbA1c Trajectory by Intervention Group",
    subtitle = "Mean ± 1 SD | 20,000 patients across 3 time points",
    x        = "Time Point",
    y        = "Mean HbA1c (%)"
  ) +
  theme_gee()
```

The close concordance between observed means (Figure 12) and model-estimated EMMs (Figure 10) confirms that the GEE mean model is not systematically over- or under-predicting for any group-time combination. Stable ±1 SD error bars across time points support the homoscedasticity assumption.

## A: Sensitivity Analysis: Cross-Structure Comparison

::: {.callout-note collapse="true"}
## Why Sensitivity Analysis?
GEE inference can be sensitive to the working correlation structure when within-patient correlation is extreme (α ≈ 0.94–0.96). Reporting key coefficients across all four structures transparently communicates structural sensitivity, recommended practice for GEE reporting.
:::

```{r sensitivity-table}
sens_tbl <- bind_rows(
  broom::tidy(gee_ind,   conf.int = TRUE) |> mutate(Structure = "Independence"),
  broom::tidy(gee_exch,  conf.int = TRUE) |> mutate(Structure = "Exchangeable"),
  broom::tidy(gee_ar1,   conf.int = TRUE) |> mutate(Structure = "AR(1)"),
  broom::tidy(gee_unstr, conf.int = TRUE) |> mutate(Structure = "Unstructured")
) |>
  filter(term %in% c("time_stepT2_Mid", "time_stepT3_Post",
                     "intervention_strengthStandard",
                     "intervention_strengthIntensive")) |>
  mutate(
    Predictor = case_match(term,
      "time_stepT2_Mid"               ~ "Time: T2_Mid",
      "time_stepT3_Post"              ~ "Time: T3_Post",
      "intervention_strengthStandard" ~ "Intervention: Standard",
      "intervention_strengthIntensive"~ "Intervention: Intensive",
      .default                        = term
    ),
    `β (95% CI)` = sprintf("%.3f (%.3f, %.3f)", estimate, conf.low, conf.high),
    `p-value` = case_when(
      p.value < 0.001 ~ "< 0.001***",
      p.value < 0.01  ~ sprintf("%.3f**", p.value),
      p.value < 0.05  ~ sprintf("%.3f*",  p.value),
      TRUE            ~ sprintf("%.3f",    p.value)
    )
  ) |>
  select(Structure, Predictor, `β (95% CI)`, `p-value`)

knitr::kable(
  sens_tbl,
  caption = "Table 7. Sensitivity Analysis - Key Predictors Across All Four GEE Working Correlation Structures"
) |>
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = TRUE
  ) |>
  pack_rows("Independence",   1, 4) |>
  pack_rows("Exchangeable",   5, 8) |>
  pack_rows("AR(1)",          9, 12) |>
  pack_rows("Unstructured",  13, 16) |>
  row_spec(c(5:8), bold = TRUE, background = "#eaf3fb") |>
  footnote(general = c(
    "Bold rows = QIC-preferred Exchangeable model (primary inference).",
    "β = population-averaged change in HbA1c (%) vs reference category.",
    "95% CI and p-values from robust sandwich variance estimator.",
    "* p < 0.05   ** p < 0.01   *** p < 0.001"
  ))
```

::: {.callout-important}
## Table 7 Interpretation: Structural Sensitivity
**Robust findings:** Time effects (T2\_Mid, T3\_Post) and Intensive intervention were significant across all four structures, these conclusions are structurally invariant. **Structure-sensitive findings:** Standard intervention was non-significant under Exchangeable (p = 0.259) and Independence, but significant under AR(1) (p = 0.022) and Unstructured (p = 0.007). The QIC-preferred Exchangeable model (highlighted rows) is the correct basis for primary inference. Standard intervention's population-level benefit is therefore statistically uncertain and should be interpreted cautiously.
:::

## P: Final Results: Population-Averaged Estimates

### Publication Table: GEE Main Effects (Exchangeable, QIC-Selected)

```{r gee-final-table}
gee_tidy <- broom::tidy(gee_exch, conf.int = TRUE) |>
  filter(term != "(Intercept)") |>
  mutate(
    Predictor   = case_match(term,
      "time_stepT2_Mid"                      ~ "Time: T2_Mid",
      "time_stepT3_Post"                     ~ "Time: T3_Post",
      "intervention_strengthStandard"        ~ "Intervention: Standard",
      "intervention_strengthIntensive"       ~ "Intervention: Intensive",
      "age"                                  ~ "Age (years)",
      "genderMale"                           ~ "Gender: Male",
      "bmi"                                  ~ "BMI (kg/m²)",
      "physical_activity_levelModerate"      ~ "Physical Activity: Moderate",
      "physical_activity_levelHigh"          ~ "Physical Activity: High",
      "diet_score"                           ~ "Diet Score",
      "sleep_hours"                          ~ "Sleep Hours",
      "medication_adherenceModerate"         ~ "Medication Adherence: Moderate",
      "medication_adherenceGood"             ~ "Medication Adherence: Good",
      "comorbidity_levelMedium"              ~ "Comorbidity: Medium",
      "comorbidity_levelHigh"                ~ "Comorbidity: High",
      .default                               = term
    ),
    `β (95% CI)` = sprintf("%.3f (%.3f, %.3f)", estimate, conf.low, conf.high),
    `p-value`    = case_when(
      p.value < 0.001 ~ "< 0.001***",
      p.value < 0.01  ~ sprintf("%.3f**", p.value),
      p.value < 0.05  ~ sprintf("%.3f*", p.value),
      TRUE            ~ sprintf("%.3f",   p.value)
    )
  ) |>
  select(Predictor, `β (95% CI)`, `Robust SE` = std.error, `p-value`)

knitr::kable(
  gee_tidy,
  digits  = 3,
  caption = "Table 8. GEE Exchangeable (QIC-Selected): Population-Averaged Estimates for HbA1c (%)"
) |>
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width        = TRUE
  ) |>
  pack_rows("Time Effects",          1, 2) |>
  pack_rows("Intervention Effects",  3, 4) |>
  pack_rows("Socio-demographic",     5, 7) |>
  pack_rows("Lifestyle Factors",     8, 11) |>
  pack_rows("Clinical Factors",     12, 15) |>
  footnote(general = c(
    "β = population-averaged change in HbA1c (%) per unit change in predictor.",
    "95% CI and p-values computed using robust sandwich variance estimator.",
    "Reference categories: Time = T1_Pre; Intervention = Weak; Gender = Female;",
    "  Physical Activity = Low; Medication Adherence = Poor; Comorbidity = Low.",
    "* p < 0.05   ** p < 0.01   *** p < 0.001",
    "QIC-preferred model: Exchangeable (QIC = 44,871.88)."
  ))
```

### GEE Multivariable Coefficient Plot

```{r fig-gee-coef, fig.cap="Figure 13. GEE multivariable coefficient plot (Exchangeable structure, QIC-preferred). Filled circles = significant (p < 0.05); open circles = non-significant. Lifestyle factors dominate effect magnitude, substantially exceeding the intervention effects."}
coef_df <- broom::tidy(gee_exch, conf.int = TRUE) |>
  filter(term != "(Intercept)") |>
  mutate(
    group = case_when(
      str_detect(term, "time_step")              ~ "Time",
      str_detect(term, "intervention_strength")  ~ "Intervention",
      term %in% c("age", "genderMale", "bmi")    ~ "Socio-demographic",
      str_detect(term, "physical_activity|diet|sleep") ~ "Lifestyle",
      str_detect(term, "medication|comorbidity") ~ "Clinical"
    ),
    label = case_match(term,
      "time_stepT2_Mid"                      ~ "Time: T2_Mid",
      "time_stepT3_Post"                     ~ "Time: T3_Post",
      "intervention_strengthStandard"        ~ "Intervention: Standard",
      "intervention_strengthIntensive"       ~ "Intervention: Intensive",
      "age"                                  ~ "Age",
      "genderMale"                           ~ "Gender: Male",
      "bmi"                                  ~ "BMI",
      "physical_activity_levelModerate"      ~ "PA: Moderate",
      "physical_activity_levelHigh"          ~ "PA: High",
      "diet_score"                           ~ "Diet Score",
      "sleep_hours"                          ~ "Sleep Hours",
      "medication_adherenceModerate"         ~ "Med Adherence: Moderate",
      "medication_adherenceGood"             ~ "Med Adherence: Good",
      "comorbidity_levelMedium"              ~ "Comorbidity: Medium",
      "comorbidity_levelHigh"                ~ "Comorbidity: High",
      .default                               = term
    ),
    significant = p.value < 0.05
  )

group_colours <- c(
  "Time"           = "#0072B2",
  "Intervention"   = "#009E73",
  "Socio-demographic" = "#E69F00",
  "Lifestyle"      = "#56B4E9",
  "Clinical"       = "#D55E00"
)

ggplot(coef_df,
       aes(x = estimate, y = reorder(label, estimate),
           colour = group, shape = significant)) +
  geom_vline(xintercept = 0, linetype = "dashed",
             colour = "#D55E00", linewidth = 0.8) +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high),
                 height = 0.3, linewidth = 0.7) +
  geom_point(size = 3) +
  scale_colour_manual(values = group_colours, name = "Predictor Group") +
  scale_shape_manual(values = c("TRUE" = 16, "FALSE" = 1),
                     name = "Significant (p < 0.05)") +
  labs(
    title    = "GEE Coefficient Plot - Exchangeable Structure (QIC-Preferred)",
    subtitle = "β estimates with 95% robust CI | Population-averaged effects",
    x        = "β (Change in HbA1c, %)",
    y        = NULL
  ) +
  theme_gee(base_size = 11) +
  theme(legend.position = "right")
```

## GEE Results Interpretation

After QIC-based selection of the **Exchangeable** working correlation structure (QIC = 44,871.88), and accounting for repeated measurements using the robust sandwich variance estimator, the following population-averaged findings emerge from the multivariable GEE analysis.

**Time effects:** Population-averaged HbA1c declined significantly over the study period. Compared to baseline (T1\_Pre), HbA1c was **−0.216%** lower at T2\_Mid (95% CI: −0.220, −0.211, p < 0.001) and **−0.420%** lower at T3\_Post (95% CI: −0.427, −0.414, p < 0.001), confirming progressive glycaemic improvement across all time points.

**Intervention effects:** Under the QIC-preferred Exchangeable model, only **Intensive** intervention achieved a significant population-averaged reduction (β = **−0.077%**, 95% CI: −0.130, −0.024, p = 0.005). Standard intervention was non-significant (β = −0.028%, p = 0.259). The sensitivity analysis showed that Standard becomes significant under AR(1) (p = 0.022) and Unstructured (p = 0.007), indicating its effect is direction ally consistent but statistically uncertain under the optimal correlation structure.

**Lifestyle factors (dominant effects):** All lifestyle predictors were independently and significantly associated with HbA1c. Physical Activity High produced the largest reduction (β = **−0.516%**, p < 0.001), followed by sleep hours (β = **−0.406%** per hour), medication adherence Good (β = **−0.355%**), and diet score (β = **−0.320%** per unit). These magnitudes substantially exceed the intervention effects, confirming that lifestyle modification is the dominant driver of population-level glycaemic control.

**Clinical factors:** Only **Comorbidity High** was significant under Exchangeable (β = +0.057%, p = 0.028), while Medium was non-significant (p = 0.272).

**Interaction:** All Time × Intervention interaction terms were highly significant in the AR(1) interaction model (all p < 0.0001). By T3\_Post, the Intensive–Weak EMM gap widened to **0.30 pp** from near-zero at baseline, demonstrating a clear dose–response temporal divergence.

## Analytical Limitations

**1. Synthetic dataset.** The dataset was sourced from Kaggle and may not represent actual Malaysian T2DM clinical heterogeneity. Findings should not be extrapolated to real-world policy decisions without validation on authentic cohort or registry data.

**2. Non-randomised allocation.** If intervention assignment was non-random, residual confounding by unmeasured variables (genetic predisposition, socioeconomic status, healthcare access) may bias the estimated intervention effects.

**3. Only three time points.** With T1\_Pre, T2\_Mid, and T3\_Post, the analysis cannot capture non-linear trajectories within the observation window or rebound patterns following initial improvement.

**4. GEE provides no individual-level predictions.** GEE estimates are population-averaged and cannot predict HbA1c for individual patients. For individual-level prediction or precision medicine applications, LMM or Bayesian hierarchical models are required.

---

## Public Health Applications

The findings from this combined LMM and GEE analysis have direct applications in Malaysian public health practice.

**Application 1: Intensive Intervention Prioritisation.** The GEE Exchangeable model identifies Intensive intervention as the only robustly significant intervention tier at the population level (β = −0.077%, p = 0.005). Combined with the LMM interaction results showing a 0.30 pp adjusted EMM gap at T3\_Post, these findings provide quantitative evidence to support **prioritisation of Intensive intervention** in the Malaysian Clinical Practice Guideline on T2DM Management (Ministry of Health Malaysia). For policy makers, GEE's population-averaged estimand is particularly appropriate for answering: "If all T2DM patients received Intensive versus Weak intervention, what is the expected population-level HbA1c difference?"

**Application 2: Multi-Component Lifestyle Modification.** Lifestyle factors dominated effect magnitudes in both models. The GEE β coefficients translate directly into expected population-level HbA1c reductions per unit change in each lifestyle domain, enabling cost-effectiveness modelling: increasing mean population physical activity from Low to High is expected to reduce population-averaged HbA1c by 0.516 percentage points, comparable to adding a full pharmacological treatment tier. Structured T2DM self-management programmes at Klinik Kesihatan should target all four lifestyle domains (physical activity, dietary quality, sleep hygiene, medication adherence) simultaneously.

**Application 3: Population-Level Glycaemic Surveillance.** The **National Diabetes Registry (NDR) Malaysia** can adopt the GEE analytical approach demonstrated here to monitor population-averaged HbA1c trajectories across clinic- or district-level cohorts over time, estimate the population-level effect of specific policy changes via time × policy interaction terms, and identify subgroup heterogeneity in glycaemic improvement across demographic strata.

---

# Conclusion {.unnumbered}

This analysis applied complementary longitudinal methods: **Linear Mixed Effects Model (LMM)** and **Generalised Estimating Equations (GEE)**, to examine HbA1c trajectories in 20,000 T2DM patients across three time points, following the REAAAPP framework throughout.

The null model ICC of **0.813** confirmed that 81.3% of total HbA1c variance is attributable to stable between-patient differences, strongly validating the need for both LMM and GEE over standard regression.

From the **LMM perspective**, all four Time × Intervention interaction terms were highly significant (p < 0.0001). Subject-specific estimates showed that the Intensive group achieved an adjusted mean HbA1c of 8.04% at T3\_Post versus 8.24% for Weak. The conditional R² of 0.974 confirmed excellent model fit. The most important modifiable predictors were high physical activity (−0.52%***), sleep hours (−0.41%***), good medication adherence (−0.36%***), and diet score (−0.32%***).

From the **GEE perspective**, the QIC-preferred Exchangeable structure (QIC = 44,871.88) confirmed that only Intensive intervention achieved a statistically significant population-averaged reduction (β = −0.077%, p = 0.005) after full covariate adjustment. Lifestyle factors dominated effect magnitudes (Physical Activity High: −0.516%; Sleep: −0.406%; Medication Adherence Good: −0.355%), substantially exceeding the intervention effects. The Time × Intervention interaction confirmed a 30-fold widening of the Intensive–Weak gap between baseline and T3\_Post.

Together, these findings support two primary policy recommendations for Malaysian T2DM management: (1) **prioritise intensive intervention protocols**, as they produce the most robust and sustained glycaemic benefit at both the individual and population level; and (2) **invest equally in multi-component lifestyle modification**, targeting physical activity, diet quality, sleep hygiene, and medication adherence, domains whose population-level effect sizes rival or exceed those of pharmacological intensification.

---

# GitHub Repository {.unnumbered}

The complete analysis code, data references, and rendered output for this assignment are available through the following repositories:

```{r github-links, echo=FALSE}
cat("Analysis Repository  : https://github.com/fizasaharudin-ux/T2DM_LMM.git\n")
cat("Analysis Repository  : https://github.com/azusa-ui/Adv-Numerical-Analysis-LMM-GEE")
```

> Both repositories contain the reproducible Quarto Markdown (`.qmd`) source files, rendered HTML outputs, and documentation for all models presented in this analysis. The combined analysis file (`T2DM_LMM_GEE_Combined.qmd`) supersedes the individual files and represents the definitive submission document.

---

# Reflection {.unnumbered}

The following section contains individual reflections from each member of the analysis team on their learning journey through this GET503 Advanced Numerical Data Analysis assignment. Each reflection addresses the methodological concepts encountered, analytical challenges resolved, and connections drawn to real-world public health practice.

---

## Dr. Siti Nurhafizah {.unnumbered}

Longitudinal data analysis is easier to understand in practice than on paper. Working through both LMM and GEE on a real T2DM dataset made the methodological distinctions concrete and, more importantly, clinically and policy-relevant. The real learning came from reconciling why both models were needed. LMM captures how individual patients change over time, which matters for clinical decisions. GEE asks something different. Across the entire population, what HbA1c reduction can we realistically expect if intensive intervention is scaled up? Those are not the same question and should not use the same method. Health directors working with program budgets need population-averaged effects, not individual trajectories. Selecting the GEE correlation structure was a decision I could not skip. QIC pointed to Exchangeable at 44,871.88, lower than AR(1) at 52,069.55. Getting there required understanding what within-patient correlation does to standard errors when independence is violated. The sandwich estimator holds inference together even when the working correlation structure is imperfect, but only if you know why you are using it. The finding that grounded everything was straightforward. By T3_Post, the Intensive group had HbA1c 0.30 percentage points lower than the Weak group. That is what a Ministry of Health proposal actually needs. I started with thinking about one patient at a time and finished thinking about what happens when a decision applies to everyone.

---

## Dr. Ahmad Zulfahmi {.unnumbered}

This assignment pushed me beyond the comfort of cross-sectional thinking. Working alongside three colleagues on a shared dataset of 20,000 T2DM patients measured at three time points, I had to genuinely reason about why standard regression fails when observations are nested within individuals, something I had previously accepted on faith rather than truly understood. The ICC of 0.813 was a striking reminder: more than 80% of HbA1c variation lived between patients, not within them. Ignoring that would have been analytically dishonest.

Building the LMM progressively, from null to interaction model, taught me that model selection is a thinking process, not a checklist. On the GEE side, fitting four correlation structures and watching QIC anoint Exchangeable over AR(1) made me appreciate that working correlation is a modelling choice with real inferential consequences, not a default setting to accept blindly.

What I found most humbling was the comparison between the two methods, and the conversations with my group made that clearer. Each of us approached the LMM and GEE outputs differently, and reconciling those perspectives deepened my understanding far beyond what I would have reached alone. LMM and GEE answer slightly different questions, subject-specific versus population-averaged, and that distinction matters enormously in public health, where policy targets populations, not individuals.

Technically, I grew. Analytically, I learned to ask "for whom is this estimate valid?" before reporting any number.

---

## Dr. Nazirul Munir {.unnumbered}

This assignment forced me to confront one dataset through two competing estimands. The same 20,000 patients, measured at three time points, analysed two ways, each demanding a different analytical logic.

The linear mixed model grounded me in an idea I had glossed over before: that repeated HbA1c readings from one patient are not independent, and that a random intercept separates within-patient change from stable between-patient differences. Seeing the null-model ICC reach 0.813, with more than four fifths of the variance sitting between patients, made the random effect feel structural rather than optional. The GEE then shifted my thinking. By marginalising over patients rather than conditioning on them, it answered a different question, and the QIC-guided choice of an exchangeable working correlation with robust sandwich errors taught me that the correlation structure is a decision, not an afterthought.

The comparison stretched me most. Both methods agreed Intensive intervention won, yet their endpoint gaps differed: the LMM gave a subject-specific contrast while the GEE gave a 0.30 percentage point population-averaged gap between Intensive and Weak at T3_Post. That surfaced a real question: whose HbA1c change am I estimating, one patient's trajectory or the population's average? That distinction decides whether a finding speaks to a clinician or to the Ministry.

---

## Dr. Fittri Fahmi {.unnumbered}

Our group initially struggled to see how LMM and GEE would translate into daily work as public health trainees, but applying both to the T2DM dataset made the relevance clear. LMM subject-specific estimates helped us understand individual trajectories, while GEE's population-averaged coefficients answered the policy question: what HbA1c reduction can we expect at scale if we roll out intensive intervention? This distinction is critical for programme evaluation. Health directors need marginal effects, not patient-level deviations. We divided tasks so one member built the LMM interaction models while another handled GEE correlation structure selection and QIC comparison. Reconciling our findings forced us to confront why the sandwich estimator matters when real-world data violate independence. The most rewarding step was deriving estimated marginal means and showing the Intensive group achieved 0.30 percentage points lower HbA1c than Weak at T3_Post. That is the quantifiable, budget-justifying evidence we need for proposing diabetes intervention intensity to the Ministry. We also learned that mis-specified reference categories or alphabetical defaults would mislead stakeholders, so clinical sense guides coding choices. Writing the QMD files pushed us to translate output into policy language. This assignment shifted our thinking from individual clinical care to population-level programme design and resource allocation.

---

# References {.unnumbered}

Hardin, J. W., & Hilbe, J. M. (2012). *Generalised estimating equations* (2nd ed.). Chapman and Hall/CRC. https://doi.org/10.1201/b13880

Højsgaard, S., Halekoh, U., & Yan, J. (2006). The R package geepack for generalised estimating equations. *Journal of Statistical Software*, *15*(2), 1–11. https://doi.org/10.18637/jss.v015.i02

Hubbard, A. E., Ahern, J., Fleischer, N. L., Van der Laan, M., Lippman, S. A., Jewell, N., Bruckner, T., & Satariano, W. A. (2010). To GEE or not to GEE: Comparing population average and mixed models for estimating the associations between neighbourhood risk factors and health. *Epidemiology*, *21*(4), 467–474. https://doi.org/10.1097/EDE.0b013e3181caeb90

Liang, K.-Y., & Zeger, S. L. (1986). Longitudinal data analysis using generalised linear models. *Biometrika*, *73*(1), 13–22. https://doi.org/10.1093/biomet/73.1.13

Nakagawa, S., & Schielzeth, H. (2013). A general and simple method for obtaining R² from generalized linear mixed-effects models. *Methods in Ecology and Evolution*, *4*(2), 133–142. https://doi.org/10.1111/j.2041-210x.2012.00261.x

Okabe, M., & Ito, K. (2008). *Color universal design (CUD): How to make figures and presentations that are friendly to colorblind people*. https://jfly.uni-koeln.de/color/

Pan, W. (2001). Akaike's information criterion in generalised estimating equations. *Biometrics*, *57*(1), 120–125. https://doi.org/10.1111/j.0006-341X.2001.00120.x

Zeger, S. L., Liang, K.-Y., & Albert, P. S. (1988). Models for longitudinal data: A generalised estimating equation approach. *Biometrics*, *44*(4), 1049–1060. https://doi.org/10.2307/2531734

---

# Acknowledgement {.unnumbered}

The dataset used in this analysis is a synthetic longitudinal cohort of 20,000 patients with Type 2 Diabetes Mellitus, measured across three time points (pre-, mid-, and post-intervention). It was sourced from Kaggle for educational and analytical practice purposes.

> **Data source:** Rana, S. *T2DM Longitudinal Treatment Outcomes* [Dataset]. Kaggle, 2024.
> Available at: [https://www.kaggle.com/datasets/sahilrana034/t2dm-longitudinal-treatment-outcomes](https://www.kaggle.com/datasets/sahilrana034/t2dm-longitudinal-treatment-outcomes)

---

*GET503 Advanced Numerical Data Analysis | DrPH Epidemiology | Siti Nurhafizah binti Saharudin · Ahmad Zulfahmi bin Sha'ari · Nazirul Munir bin Abu Hassan · Mohd Fittri Fahmi bin Fauzi | `r format(Sys.Date(), "%B %Y")`*
